---
title: "crossvalidation vs MAP"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan)
library(glmnet)
```

# Crossvalidation vs MAP
Here we generate correlated data

$$ height \sim N(10, 2) \\
legprop \sim uniform(Nsamples, 0.4, 0.5)$$

```{r data generation}
N_samples <- 1000
height <- rnorm(N_samples, 10, 2)
leg_prop <- runif(N_samples, 0.4, 0.5)
leg_left <- leg_prop * height + rnorm(N_samples, 0, 0.02)
leg_right <- leg_prop * height + rnorm(N_samples, 0, 0.02)
d <- data.frame(height, leg_left, leg_right)
```


```{stan output.var="height_inference_6_1"}
data {
  int<lower=0> N_samples;
  real<lower=0> b_std;
  vector[N_samples] height;
  vector[N_samples] leg_left;
  vector[N_samples] leg_right;
}
parameters {
  real a;
  real bl;
  real br;
  real<lower=0> sigma;
}
model {
	a ~ normal(10, 100);
	bl ~ normal(2, b_std);
	br ~ normal(2, b_std);
  sigma ~ uniform(0,10);
  height ~ normal(a + bl *leg_left + br * leg_right, sigma);
}
```
## MAP estimate for estimating mu at max sample mean

```{r}
b_std <- 10
fit <- optimizing(height_inference_6_1, data = list(b_std=b_std, N_samples=N_samples, height=height, leg_left=leg_left, leg_right=leg_right), as_vector=FALSE, hessian=TRUE)
fit$par
vcov <-  solve(-fit$hessian)
print(vcov)

sds <- sqrt(diag(vcov))
lambda <- (fit$par$sigma/sds["bl"])^2/N_samples
#?? vcov 

```

Use penalised ML with glmnet.
the regularisation defaults much stronger.  How is this identified?
(but then crossvalidate) .. and might not identify minimum in section

Cannot set this up as a random effects model?

```{r}

X <- as.matrix(d[,c("leg_left", "leg_right")])
y <- as.matrix(d$height)
md <- cv.glmnet(X, y, alpha=0, standardize=TRUE, lambda = seq(0.0001,0.1,length.out=1000))
```

```{r}
cat(md$lambda.min, md$lambda.1se, lambda, fit$par$sigma)
sigma_b <- fit$par$sigma / sqrt(md$lambda.min *N_samples)
fit$par
print("coefficient using MAP lambda estimate")
coef(md, lambda)
print("coefficient based on crossvalidation")
coef(md, md$lambda.min)

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
